---
title: "An optimization approach to separate the latent deterioration process from maintenance intervention on navyship failure engineering"
output: pdf_document
---

This is a supplementary notebook to the original paper.

```{r, results='hide'}
library(ggplot2)
library(tidyverse)
library(rstan)
library(scales)
library(knitr)
```

## Exploratory Data Analysis

Our target data consist of annual failure counts per vessel. Each data point consists of a set of the following:

* Annual Failure Count
* Age
* Ship Type indentifier
* Engine Type indentifier

Data accumulation started from the initial recording year for 31 years, with a total of 99 individual ships; each containing an engine from 4 different types.


```{r}
data <- read.csv("failure_count.csv")[, 2:100]
data2 <- data %>% 
  rownames_to_column(var = "age") %>% 
  gather(key="ship", value="value", -age) %>%
  mutate(ship=as.integer(substring(ship, 2)), age=as.integer(age), value=as.integer(value))

head(data2)
```
```{r}
ggplot(data2, aes(x = ship, y = age, fill=value)) + 
  scale_fill_gradientn(colours = rev(rainbow(3)), na.value = "white") + geom_tile() + 
  scale_y_reverse() + 
  coord_equal()
```

## Data Imputation

As you can see from the above heatmap, portions of the data are missing. Since individual ships shared common traits if they are of the same ship/engine type,
we deduced that such information may be approximated by combining per ship and per engine data.

The failure count process was modeled as a Gaussian process, with per ship and engine parameters for mean and variance.
This was implemented in stan with a exponentiated quadratic kernel for covariance parameters in each Gaussian process:
```{stan output.var="impute_gp_model", eval = FALSE}
data {
  int<lower=1> N; 
  int<lower=1> N_engines; 
  int<lower=1> N_ships;
  int<lower=1> N_ages_obs;
  int<lower=1> N_ages;
  int<lower=1> ship_engine_ind[N_ships];
  int<lower=1,upper=99> ship_ind[N];
  int<lower=1> age_ind[N]; 
  vector[N] y; 
  real<lower=0> hp_scale;
  real<lower=0> emp_le_shape;
  real<lower=0> emp_le_scale;
  real<lower=0> emp_ls_shape;
  real<lower=0> emp_ls_scale;
}

transformed data {
  real ages[N_ages];
  int N_comp = 10;
  for (t in 1:N_ages)
    ages[t] = t;
}

parameters {
  matrix[N_ages,N_engines] GP_engine_std;
  matrix[N_ages,N_ships] GP_ship_std;
  vector[N_ages_obs] age_std;
  vector[N_ships] ship_std;
  vector[N_engines] engine_std;
  real<lower=0> tot_var;
  simplex[N_comp] prop_var;
  real mu;
  real<lower=0> length_GP_engine_s;
  real<lower=0> length_GP_ship_s;
  real <lower = 0> length_engine_scale;
  real <lower = 0> length_ship_scale; 
  real <lower = 0> length_engine_shape;
  real <lower = 0> length_ship_shape;
}

transformed parameters {
  matrix[N_ages,N_engines] GP_engine;
  matrix[N_ages,N_ships] GP_ship;

  vector[N_ages_obs] age_re;
  vector[N_ships] ship_re;
  vector[N_engines] engine_re;
  vector[N_comp] vars;
  
  real sigma_age;
  real sigma_engine;
  real sigma_ship; 

  real sigma_error_ship[5];

  real sigma_GP_engine;
  real sigma_GP_ship;
  
  real length_GP_engine = length_engine_scale * length_GP_engine_s;
  real length_GP_ship = length_ship_scale * length_GP_ship_s;
  
  vars = N_comp * prop_var * tot_var;
  sigma_age = sqrt(vars[1]);
  sigma_engine = sqrt(vars[2]);
  sigma_ship = sqrt(vars[3]); 
  sigma_GP_engine = sqrt(vars[4]);
  sigma_GP_ship = sqrt(vars[5]);
  for (i in 6:N_comp){
      sigma_error_ship[i-5] = sqrt(vars[i]);
  }

  engine_re = sigma_engine * engine_std;
  age_re = sigma_age * age_std;
  ship_re = sigma_ship * ship_std; 
  
  {
    matrix[N_ages, N_ages] cov_engine; 
    matrix[N_ages, N_ages] cov_ship; 
    matrix[N_ages, N_ages] L_cov_engine; 
    matrix[N_ages, N_ages] L_cov_ship; 

    cov_engine = cov_exp_quad(ages, sigma_GP_engine, 
                                  length_GP_engine);
    cov_ship = cov_exp_quad(ages, sigma_GP_ship, 
                                  length_GP_ship);
    for (age in 1:N_ages) {
      cov_engine[age, age] = cov_engine[age, age] + 1e-6;
      cov_ship[age, age] = cov_ship[age, age] + 1e-6;
    }

    L_cov_engine = cholesky_decompose(cov_engine);
    L_cov_ship = cholesky_decompose(cov_ship);
    GP_engine = L_cov_engine * GP_engine_std; //f_engine
    GP_ship = L_cov_ship * GP_ship_std;       //f_ship
  }
}

model {
  vector[N] obs_mu;
  for (n in 1:N) {
    obs_mu[n] = mu 
              + age_re[age_ind[n]]                                 //fixed effects
              + engine_re[ship_engine_ind[ship_ind[n]]] 
              + ship_re[ship_ind[n]]   
              + GP_engine[age_ind[n],ship_engine_ind[ship_ind[n]]] //f_engine 
              + GP_ship[age_ind[n],ship_ind[n]];                   //f_ship
    y[n] ~ normal(obs_mu[n], sigma_error_ship[ship_engine_ind[ship_ind[n]]]);
  }

  to_vector(GP_engine_std) ~ normal(0, 1);
  to_vector(GP_ship_std) ~ normal(0, 1);
  age_std ~ normal(0, 1);
  ship_std ~ normal(0, 1);
  engine_std ~ normal(0, 1);
  mu ~ normal(.5, .5);
  tot_var ~ normal(0,1);
  length_engine_shape ~  normal(emp_le_shape, hp_scale);
  length_engine_scale ~ normal(emp_le_scale, hp_scale);
  length_ship_shape ~  normal(emp_ls_shape, hp_scale);
  length_ship_scale ~ normal(emp_ls_scale, hp_scale);
  length_GP_engine_s ~ weibull(length_engine_shape,1);
  length_GP_ship_s ~ weibull(length_ship_shape,1);
}
```

Note that the imputed data was passed through a Yeo-Johnson transform.

```{r}

imputed_data_gp <- read.csv("y_pred_5var.csv")[,-1]
imputed_data <- imputed_data_gp %>% 
  rownames_to_column(var = "age") %>% 
  gather(key="ship", value="y_data", -age) %>%
  mutate(ship=as.integer(substring(ship, 2)), age=as.integer(age), y_data=as.numeric(y_data))
ggplot(imputed_data, aes(x = ship, y = age, fill=y_data)) + 
  scale_fill_gradientn(colours = rev(rainbow(3))) + geom_tile() + 
  scale_y_reverse() + 
  coord_equal()
```

## Deterioration and Maintenance Model

We took an approach to model the entire degradation process as a Markov Chain for ease of interpretability.

Annual failure counts were subdivided into 3 quantiles, each given a singer state within the Markov chain.

We had to infer both the positive effects of maintenance and the speed of deterioration, which are both present within the data.
Maintenance is also done every year, with varying effects.

The overall transition rate matrix was split into 2 matrices, D and M, which modeled respectively the effects of deterioration and maintenance.
Both were estimated from the data by applying both matrices to each Markov transition, selecting the matrices which yielded the lowest error.
## Age changepoints

In order to take into account the potential factor of age into the rate of deterioration, we decided to estimate a deterioration matrix per for each of 3 age subgroups.

Our initial attempt was to use Stan's pointwise optimizer and find the time changepoints by minimizing the squared error.

Unfortunately, working with integer parameters, even the marginalization trick was insufficient in generating a stable solution.

We decided to resort to `envcpt` which includes a time series changepoint detection model. 
The resulting changepoint ages were 8, 20, and 26.

```{r}
# insert changepoint code here
```

## Modeling the transition rate matrices
### Deterioration rate matrix
Once the quantiles and age changepoints were decided, a parameterization for the transition rate matrix is required.

Since deterioration can only occur on a forward manner, for a system with 3 states, 3 parameters, each representing the transition rates to subsequent states are required.

Since transition rates are dependent on time, we need to calculate the deterioration matrix for each time t. Fortunately a closed form solution exists given all parameters are known:

(insert matrix equation here)

￩
### Maintenance rate matrix

The maintenance model for repair ship engines are defined as follows: for states 2, 3 maintenance may transition from the current state to any of the forward states based on a repair effectiveness probability, or even have not effect and stay at the current state. Given repair probability p and q, the maintenance rate matrix can be parameterized as follows:

(insert p repair probability matrix equation here)

## Combining rate matrices and estimating the matrix parameters

The deterioration and maintenance matrix combined make up the joint transition rate matrix, which was estimated with failure data through stan's pointwise optimizer:

```{r}
stanmodel_code <- "
data {
  int<lower=0> N; // number of observations
  int<lower=1>n_state; // number of states
  vector[n_state] state_obs[N];
  int time_obs[N];
  int<lower=0, upper=n_state> initial_state;
}

transformed data {
  vector[n_state] initial;
  for(i in 1:n_state){
    initial[i] = 0;
    if(i == initial_state){
      initial[i] = 1;
    }
  }
}

parameters {
  real<lower=0> rate[4,3];
  real<lower=0, upper=1> p;
  real<lower=0, upper=1> q;
}

transformed parameters {
  matrix[n_state, n_state] DM_pow[31];
  matrix[n_state, n_state] D[4];
  matrix <lower =0>[n_state, n_state] M;
  
  M[1,1]=1;
  M[1,2]=p;
  M[1,3]=q;
  M[2,1]=0;
  M[2,2]=(1-p);
  M[2,3]= (1-q);
  M[3,1]=0;
  M[3,2]=0;
  M[3,3]= 0; //(1-q-r);
  
  for(i in 1:4){
    D[i][1,1] = exp(-(rate[i,1]+ rate[i,2]));
    D[i][2,1] = rate[i,1] * exp(-rate[i,3]) * (1-exp(-(rate[i,1]+ rate[i,2] - rate[i,3]))) / (rate[i,1]+ rate[i,2] - rate[i,3]);
    D[i][1,2] = 0;
    D[i][3,1] = 1 - D[i][1,1] - D[i][2,1];
    D[i][2,2] = exp(-rate[i,3]);
    D[i][3,2] = 1 - D[i][2,2];
    D[i][1,3] = 0;
    D[i][2,3] = 0;
    D[i][3,3] = 1;
  }
  
  DM_pow[1] = D[1];
  for (i in 2:max(time_obs)){
    if (i <= 8){
      DM_pow[i] = D[1] * M * DM_pow[i-1];
    }
    else if (i <=20){
      DM_pow[i] = D[2] * M * DM_pow[i-1];
    }
    else if (i <=26){
      DM_pow[i] = D[3] * M * DM_pow[i-1];
    }
    else{
      DM_pow[i] = D[4] * M * DM_pow[i-1];
    }
  }

}

model {

  for(i in 1:N){
    target += -(DM_pow[time_obs[i]]  * initial - state_obs[i])'*(DM_pow[time_obs[i]] * initial - state_obs[i]);
  }
}

"
```

The data is divided into 3 subgroups depending on quantiles and one-hot encoded:

```{r}
model_DMsep = stan_model(model_code = stanmodel_code, verbose = FALSE)

n_state = 3
initial_state = 1

generate_state_matrix <- function(data, n){
  #state <- cut(data, breaks=c(0, 80, 160, max(data)), labels=1:n, include.lowest = TRUE)
  state <- cut(data, breaks=quantile(data,c(0,1/3,2/3,1)), labels=1:n, include.lowest = TRUE)
  state<-as.numeric(state)
  matrix(state,nrow=31)
}

state_matrix <- generate_state_matrix(imputed_data$y_data, n_state)
states <- as.vector(t(state_matrix)) #[imputed_data$engine_ind == engine_type] shiptype -> year
onehot <- list()
# one-hot encode per-data state to vector
for(i in 1:length(states)){
  t_tmp <- rep(0, len=n_state)
  t_tmp[states[i]] <- 1
  onehot[[i]] <- t_tmp
}
onehot_array <- aperm(array(unlist(onehot),c(n_state, length(onehot)))) # array() fills column-wise first
head(onehot_array)
```

Once the data and model are ready, we run optimization 1000 individual times:

```{r R.options=list(max.print=20)}
iter=100

MSE_df<-data.frame(index=rep(0,iter),test_MSE=rep(0,iter),p=rep(0,iter),q=rep(0,iter),train_MSE=rep(0,iter),rate1=rep(0,iter),rate2=rep(0,iter),rate3=rep(0,iter))
D_array <- array(0, dim=c(iter,4,3,3))
ship_ind_df<-matrix(0,nrow=iter,ncol=5)

for (i in 1:iter){
  test_ship_ind= sort(sample(1:99,5)) #c(17,20,24,77,82)
  ship_ind_df[i,]=test_ship_ind
  test_ind=c(sapply(test_ship_ind,function(x) (x-1)*31+(1:31)))
  train_data <- onehot_array[-test_ind, ]
  test_data <- onehot_array[test_ind, ]
  opt_data <- list(N= dim(train_data)[1], n_state=n_state, state_obs=train_data, time_obs=imputed_data$age[-test_ind], initial_state=initial_state)
  
  res <- optimizing(model_DMsep, opt_data, iter = 2000, verbose = TRUE,hessian = TRUE, history_size=10, init = list(rate=array(c(0.5,0.5,0.5,0.5,0.1,0.1,0.1,0.1,0.5,0.5,0.5,0.5), dim = c(4, 3))))
  
  predicted_state<-matrix(0,nrow=31,ncol=3)
  DM_pow<-array(0,dim=c(31,3,3))
  
  for (t in 1:31){
    for (k in 1:3){
      for (j in 1:3){
        DM_pow[t,k,j]=res$par[paste0("DM_pow[",t,",",k,",", j,"]")]
      }
    }
    predicted_state[t,]=DM_pow[t,,]%*%c(1,0,0)
  }
  MSE_df[i,3]=res$par["p"]
  MSE_df[i,4]=res$par["q"]
  
  for (era in 1:4){
    D <- matrix(as.vector(unlist(lapply(1:n_state, function(row){lapply(1:n_state, function(col){res$par[paste0("D[",era,",",row,",",col,"]")]})}))), nrow=3, byrow=T)
    D_array[i,era,,]=D
  }
  MSE_df[i,6]=res$par["rate[3,1]"]
  MSE_df[i,7]=res$par["rate[3,2]"]
  MSE_df[i,8]=res$par["rate[3,3]"]
  
  SSE_total = rep(0,nrow=99*31)
  for (ind in 1:(99*31)){
    SSE_total[ind]=sum((onehot_array[ind,]-predicted_state[(ind-1)%%31+1,])^2)
  }
  MSE_df[i,1]=i
  MSE_df[i,2]=sum(SSE_total[test_ind])/5
  MSE_df[i,5]=sum(SSE_total[-test_ind])/94
}

```
Plot the distribution of train and test MSE for each iteration:

```{r}
par(mfrow=c(1,1))
hist(MSE_df$test_MSE,breaks=100,main="test MSE",xlab="test MSE")
hist(MSE_df$train_MSE,breaks=100,main="train MSE",xlab="train MSE")
```

And...the approximated transition rate matrices for each time interval:

```{r}
rate<-matrix(0,nrow=4,ncol=3)

for (era in 1:4){
  D <- matrix(as.vector(unlist(lapply(1:n_state, function(row){lapply(1:n_state, function(col){res$par[paste0("D[",era,",",row,",",col,"]")]})}))), nrow=3, byrow=T)
  for(j in 1:3){
    rate[era,j] <- res$par[paste0("rate[", era,",", j,"]")]
  }
  #h_ <- hist(rate[era,], breaks=50)
  #print(h_$counts * h_$breaks)
  print(paste0("Era:",era))
  print(D)
}
```
Finally, we'll plot the states predicted by the model:

```{r}
observed_count = sapply(1:3,function(i) apply(state_matrix,1,function(x) sum(x==i)))
observed_prop=t(apply(observed_count,1,function(x) x/sum(x)))
observed_scaled_state = exp(exp(exp(observed_prop)))/10
observed_scaled_state = t(apply(observed_scaled_state,1,function(x) x/sum(x)))
scaled_state=exp(exp(exp(predicted_state)))/10
scaled_state =  t(apply(scaled_state,1,function(x) x/sum(x)))

ship_observed_state = data.frame(ship_id=rep(1:99,31),time=factor(rep(1:31,each=99)),state=states)
total_count_df=data.frame(time=factor(rep(1:31,3)),states=factor(rep(1:3,each=31)),observed=c(observed_count),predicted=c(predicted_state)*99)
test_count_df = data.frame(time=rep(1:31,5),observed=c(state_matrix[,test_ship_ind]),ship_ind=factor(rep(test_ship_ind,each=31)),predicted_max=rep(apply(predicted_state,1,which.max),5))

legend_size = 5

# ggplot(total_count_df, aes(x = time, y = states)) +
# geom_point(aes(colour="observed", size = observed),alpha=1) +
# geom_point(aes(colour="predicted", size = predicted),alpha=0.4) +
# scale_size(breaks = c(-3:3), range = c(1,30)) +
# guides(
#   size=guide_legend(override.aes = list(size = legend_size))
# ) 

total_df=data.frame(time=rep(1:31,3),states=rep(1:3,each=31),observed=c(observed_count),predicted=c(predicted_state)*99)

ggplot(test_count_df, aes(x = time, y = observed)) +
geom_line(aes(colour=ship_ind),size=1.2) +
coord_cartesian(ylim = c(0.5, 3.5)) +
geom_point(aes(colour=ship_ind),size=3) +
geom_point(data=total_df, aes(x=time,y=states, size = predicted),color="red",alpha=0.1) +
scale_size(breaks = c(-3:3), range = c(1,30)) +
guides(
  size=guide_legend(override.aes = list(size = legend_size))
) 

# ggplot(test_count_df, aes(x = time, y = observed)) +
# geom_line(aes(colour=ship_ind),size=1.2) +
# coord_cartesian(ylim = c(0.5, 3.5)) +
# geom_point(aes(colour=ship_ind),size=3) +
# geom_point(aes(x=time,y=predicted_max),size=10,color="red",alpha=0.2)
```
